---
title: "`WGCNA Analysis For `r params$projectName``"
author: "Raymond Yuan SHANG"
date: "`r params$date`"
output:
  html_document:
    standalone: TRUE
    embed-resources: TRUE
    df_print: paged
always_allow_html: yes
params:
  projectName: ROSMAP
  date: 10/08/2023
  sampleInfo: sampleInfo.xlsx
  dataDir: ~/Dropbox/human/ROSMAP/RNASeq
  groupName: NCI
  groups: ["NCI.F.APOE33","NCI.F.APOE23","NCI.M.APOE33", "NCI.M.APOE34","NCI.M.APOE23","NCI.F.APOE34"]
  powerEstimate: 0
  thresholdZ: -2.5
  MEDissThres: 0.4
  maptype: signed
  outputDir: "."
  wdir: "."

fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
#knitr::opts_chunk$set(root.dir = wdir)

#opts_knit$set(root.dir = 'c:/Users/kwilliams/git/ProjectX')
#knitr::opts_knit$set(root.dir = '/tmp')
#knitr::opts_knit$get(root.dir)
#setwd(wdir)
#getwd()
#getwd()
#Settings
options(stringsAsFactors=FALSE)
options(warn=-1)
options(DT.options = list(columnDefs = list(list(className = 'dt-center', targets = "_all"))))

```


```{r}
#rm(list=ls())
suppressPackageStartupMessages({
# Load R packages required
# For data processing
library("readxl")
library("writexl")
library("openxlsx")

  # For plotting
library("ggplot2")
library("gridExtra")
library("corrplot")
library("RColorBrewer")
library("pheatmap")
library("gplots")
library("ggh4x")
## For PCA
library("FactoMineR")
#library("")
# For stringr wrangling
library("stringr")
library("tidyverse")
# For GSEA analysis and visualization
library("msigdbr")
library("org.Hs.eg.db")
library("clusterProfiler")
library("ggstance")
# For rendering
library("htmltools")
library("DT")
  
# Others
library("sva") #In case to remove unknown batch effects using SVA
library("vsn")
library("hexbin")
library("factoextra")
  
#WGCNA
library("WGCNA")
library("cluster")
library("flashClust")
})

```

```{r}
# Check output dir
outputDir <- params$outputDir
dataFile <- params$dataFile
projectName <- params$projectName

dataDir <- params$dataDir
wdir <- params$wdir
powerEstimate <- params$powerEstimate
thresholdZ.k <- params$thresholdZ
MEDissThres <- params$MEDissThres
maptype <- params$maptype
current_groupname <- params$groupName
current_groups <- params$groups
```

```{r,include=FALSE}
#Set plotting THeme, general
theme_set(theme_bw()) + theme_update(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),plot.title = element_text(hjust = 0.5),legend.title=element_blank()) #legend.position = c(0.8, 0.8),
```


```{r}
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = "\n")
}
```



```{r}
## set dynamic render fig size in chunk
figNum=1
subchunkify <- function(g, fig_height=7, fig_width=5,chunk_name="plot") {
  g_deparsed <- paste0(deparse(
    function() {g}
  ), collapse = '')
  chunk_name <- paste0(chunk_name,".",figNum)
  sub_chunk <- paste0("\n`","``{r ",chunk_name, 
                      ", fig.height=", fig_height, 
                      ", fig.width=", fig_width, 
                      ", echo=FALSE}",
                      "\n(", 
                        g_deparsed
                        , ")()",
                      "\n`","``\n")
  figNum <<- figNum+1
  if(figNum > 100000){figNum<<-1}
  cat(knitr::knit(text = knitr::knit_expand(text = sub_chunk), quiet = TRUE))
}

```

```{r, include=FALSE}
#Magic:https://github.com/rstudio/DT/issues/67#issuecomment-426329442
#To display DT,figs in loop
#library(DT)

#datatable(iris)
DT::datatable(matrix())

```

```{r}

wp2gene_GO <- msigdbr(species = "Homo sapiens", category = "C5") %>% #,subcategory = "GO:BP"
  dplyr::select(gs_name, gene_symbol)%>% dplyr::distinct(gs_name, gene_symbol)
colnames(wp2gene_GO)=c("term","gene")
```


```{r}
#Test
dataFile = file.path (dataDir,"lcpm_WithoutFactors_ROSMAP.Rdata")#"~/Dropbox/human/ROSMAP/RNASeq/lcpm_WithoutFactors_ROSMAP.Rdata"
s2cFile = file.path(dataDir,"x_normalized_ROSMAP.Rdata")
```

```{r}

#save(lcpm_WithoutFactors,file="~/Dropbox/human/ROSMAP/RNASeq/lcpm_WithoutFactors.Rdata")
#save(y,file="~/Dropbox/human/ROSMAP/RNASeq/y.Rdata")
current_area="ROSMAP" #
#current_groupname="NCI"
#current_groups=c("NCI.F.APOE33","NCI.F.APOE23","NCI.M.APOE33", "NCI.M.APOE34","NCI.M.APOE23","NCI.F.APOE34")
#c("NCI.F.APOE33","NCI.F.APOE23","NCI.M.APOE33", "NCI.M.APOE34","NCI.M.APOE23","NCI.F.APOE34") #NCI
#c("MCI.F.APOE33","MCI.F.APOE23","MCI.M.APOE33", "MCI.M.APOE34","MCI.M.APOE23","MCI.F.APOE34") #MCI
#c("AD.F.APOE33","AD.F.APOE23","AD.M.APOE33", "AD.M.APOE34","AD.M.APOE23","AD.F.APOE34") #AD
#c()

# [1] NCI.F.APOE33 NCI.F.APOE23 MCI.F.APOE33 NCI.M.APOE33 NCI.M.APOE34 AD.M.APOE33  AD.F.APOE34  AD.M.APOE34  MCI.M.APOE33 NCI.M.APOE23
#[11] MCI.M.APOE34 MCI.M.APOE23 AD.F.APOE33  MCI.F.APOE23 MCI.F.APOE34 AD.M.APOE23  AD.F.APOE23  NCI.F.APOE34
#current_area="TCX"

load(dataFile)
load(s2cFile)
genes_detected_ROSMAP <- rownames(lcpm_WithoutFactors)
s2c <- y$samples
rownames(s2c) <- paste0("X",s2c$rna_id)
sum(rownames(s2c) != colnames(lcpm_WithoutFactors)) #Should be 0
sum(y$genes$ENSEMBL != genes_detected_ROSMAP) #should be 0
rownames(lcpm_WithoutFactors) <- y$genes$HGNC_SYMBOL
genes_detected_ROSMAP <- rownames(lcpm_WithoutFactors)

```

# WGCNA network analysis

```{r}
#Transpose expression data to WGCNA format, genes as collumns and samples as rows.
s2c_WGCNA <- s2c[s2c$groupCom %in% current_groups,]

data_WGCNA <- as.data.frame(t(lcpm_WithoutFactors[,rownames(s2c_WGCNA)]))

```


## 1. Clustering data based on distance, to check potential outliers
```{r}
# sample network based on squared Euclidean distance
# note that we transpose the data

A=adjacency(t(data_WGCNA),type="distance")
# this calculates the whole network connectivity
k=as.numeric(apply(A,2,sum))-1
# standardized connectivity
Z.k=scale(k)

# Designate samples as outlying
# if their Z.k value is below the threshold
thresholdZ.k=-3 # often -2.5
print(paste0("threhold for outliers:",thresholdZ.k))
# the color vector indicates outlyingness (red)
outlierColor=ifelse(Z.k<thresholdZ.k,"red","black")
# calculate the cluster tree using flahsClust or hclust
sampleTree = flashClust(as.dist(1-A), method = "average")

#plot(sampleTree)
datColors=data.frame(outlierC=outlierColor)
plotDendroAndColors(sampleTree,groupLabels=names(datColors),colors=datColors,main="Sample dendrogram and trait heatmap")

```


```{r}
outliers <- sampleTree$labels[datColors$outlierC=="red"]
data_WGCNA <- data_WGCNA[!(rownames(data_WGCNA) %in% outliers),]
s2c_WGCNA <- s2c_WGCNA[!(rownames(s2c_WGCNA) %in% outliers),]

```


## 2. Selection soft threshold beta value via scale free topology, for the co-expression network

```{r}
# Choose a set of soft thresholding powers
powers=c(1:25) # in practice this should include powers up to 20.
# choose power based on SFT criterion
sft=pickSoftThreshold(data_WGCNA,powerVector=powers,corFnc="bicor",networkType = maptype)
#sft
```

### 2.1 Plot the SFT results:
```{r add.layer,  echo=-1}
#Digression: if you want to pick a soft threshold for a signed network write
#sft=pickSoftThreshold(data_ROSMAPmt_NCI2,powerVector=powers, networkType = "signed")
# but then you should consider higher powers. Default beta=12.
names(sft)
sft["fitIndices"]
# Plot the results:
{
par(mfrow=c(1,2))
# SFT index as a function of different powers
plot(sft$fitIndices[,1],-sign(sft$fitIndices[,3])*sft$fitIndices[,2],xlab="Soft Threshold (power)",ylab=paste0("SFT, ",maptype," R^2"),type="n",main=paste("Scale independence"))
text(sft$fitIndices[,1],-sign(sft$fitIndices[,3])*sft$fitIndices[,2],labels=powers,col="red")
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of different powers
plot(sft$fitIndices[1:10,1],sft$fitIndices[1:10,5],type="n",
xlab="Soft Threshold (power)",ylab="Mean Connectivity",main=paste("Mean connectivity"))
text(sft$fitIndices[,1],sft$fitIndices[,5],labels=powers,col="red")
}
```


### 2.2 Check β value, whether it fits the scale free assumption
```{r}
if(powerEstimate ==0){
  softPower = sft$powerEstimate
}else{
  softPower = powerEstimate
}
#softPower=15  #sft$powerEstimate 15 for NCI,

#ADJ1_cor <- abs(WGCNA::cor(data_hypo,use = "p" ))^softPower
# Genes（<5000）, use the code below：
#k <- as.vector(apply(ADJ1_cor,2,sum,na.rm=T))
# Genes（>5000）, use the code below
k <- softConnectivity(datE=data_WGCNA,power=softPower) 
{
sizeGrWindow(10, 5)
par(mfrow=c(1,2))
hist(k)
scaleFreePlot(k,main="Check Scale free topology\n")
}
```

## 3. Adjacent Matrix, Module detection based on (1-TOM) and auto dynamic tree cutting

### 3.1 Automatic module detection via dynamic tree cutting
```{r}
clusterMethod = "average"
##  The following Matrix adj are too big (~8G) and can not run in local PC. Need to find some otherway
adj= adjacency(data_WGCNA,type = maptype, power = softPower,corFnc="bicor");#signed,unsigned
#turn adjacency matrix into topological overlap to minimize the effects of noise and spurious associations
TOM=TOMsimilarity(adj,TOMType = maptype); #,networkType = "unsigned"

#Calculating TOM similarity from Expr data directly could save memory because we only need TOM now. still it is ~8G
#TOM=TOMsimilarityFromExpr(data_hypo,networkType = "unsigned",TOMType = "signed", power = softPower);
#dim(gene_id)
gene_symbol <- genes_detected_ROSMAP
dim(gene_symbol)
#names(gene_id)
names(gene_symbol)
dim(TOM)#15582 15582

length(gene_symbol)
rownames(TOM) = gene_symbol
colnames(TOM) = gene_symbol
dissTOM=1-TOM
#hierarchical clustering of the genes based on the TOM dissimilarity measure
geneTree = flashClust(as.dist(dissTOM),method=clusterMethod); #average
#hierTOM = hclust(as.dist(dissTOM),method="average");
#plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="",cex=0.3);
```



### Gene dendrogram and module colors before merge

```{r}
# Dynamic cuttings：
minModuleSize = 30;
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                            deepSplit = 2, pamRespectsDendro = FALSE,
                            minClusterSize = minModuleSize);
#table(dynamicMods)
dynamicColors = labels2colors(dynamicMods)
#table(dynamicColors)
#geneTree$labels[dynamicColors=="tan"]
```


```{r}
# 使用相异度来聚类为gene tree(聚类树)：
#geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
#windows()
# Plot the dendrogram and colors underneath
getwd()
{
  #file.path(outputDir,paste0("Gene dendrogram and module colors before merge (",current_groupname,"_",current_area,").pdf"))
pdf(file.path(outputDir,paste0("Gene dendrogram and module colors before merge (",current_groupname,"_",current_area,").pdf")), height=10, width=15)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors before merge")
dev.off()
}

```

### 3.2 Calculate the eigen gene for each module, which represents the whole module expression level
```{r}
#lapply(data_hypo,as.numeric,2) -> data_hypo
#as.data.frame(data_hypo) -> data_hypo
MEList = moduleEigengenes((as.matrix(data_WGCNA)), colors = dynamicColors,softPower=softPower)
MEs = MEList$eigengenes
# 计算根据模块特征向量基因计算模块相异度：
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "complete")
#METree = flashClust(as.dist(MEDiss), method = "average")
# Plot the result

plotEigengeneNetworks(MEs, 
                      "Eigengene adjacency heatmap", 
                      marHeatmap = c(3,4,2,2), 
                      plotDendrograms = FALSE, 
                      xLabelsAngle = 90) 
```

Module Clustering Trees based on module eigen values
```{r}
plot(METree, 
     main = "Clustering of module eigengenes",
     xlab = "", 
     sub = "")
```

### 3.3 Save module genes
```{r}
colors_mod <- unique(dynamicColors)
df_mod <- data.frame(gene=NA,modulecolor=NA)
for(i in c(1:length(colors_mod))){
  genes_currentmodule <- geneTree$labels[dynamicColors==colors_mod[i]]
  df_mod <- rbind(df_mod,data.frame(gene=genes_currentmodule,modulecolor=colors_mod[i]))
}
#geneTree$labels[dynamicColors=="tan"]
df_mod <- df_mod[2:dim(df_mod)[1],]
write.csv(x=df_mod,file=file.path(outputDir,paste0("modules_",maptype,"_unmerged_functionAnnotation",current_area,"_",current_groupname,".csv")), row.names = FALSE)
head(df_mod)
```

### 3.3 Select 400 random genes for the toppology overlap heatmap, deep yellow and red means highly topology overlapped。
```{r}
nGenes = length(gene_symbol)
# toppology overlap heatmap：
nSelect = 400 
# For reproducibility, we set the random seed 
set.seed(10); 
select = sample(nGenes, size = nSelect); 
selectTOM = dissTOM[select, select]; 
# There's no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster. 
selectTree = hclust(as.dist(selectTOM), method = clusterMethod)#average 
selectColors = dynamicColors[select]; 
# Open a graphical window 
sizeGrWindow(9,9) 
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing 
# the color palette; setting the diagonal to NA also improves the clarity of the plot 
plotDiss = selectTOM^softPower; 
diag(plotDiss) = NA; 
{
  pdf(file.path(outputDir,paste0("Gene dendrogram and heatmap before merge (",current_groupname,"_",current_area,").pdf")), height=10, width=15)

  TOMplot(plotDiss, 
        selectTree, 
        selectColors, 
        main = "Network heatmap plot, selected genes") 
  dev.off()
}
```



## 4. Merge close modules

### 4.1 set the MEDissThres as the threshold based on the cluster tree

```{r}
pdf(file.path(outputDir,paste0("Clustering of module eigengene before merge_",current_groupname,".pdf")), height=10, width=15)

{
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
# We choose a height cut of 0.2, corresponding to correlation of 0.8
MEDissThres = 0.4


# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
}
dev.off()

```

### 4.2 New Merged Modules
```{r}
# Draw a cutting line in the cluster figure
#abline(h=MEDissThres, col = "red")
# merge：
merge_modules = mergeCloseModules(data_WGCNA, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# new colors：
mergedColors = merge_modules$colors;
# new module eigen values：
mergedMEs = merge_modules$newMEs;
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)


```

Save the plot
```{r}
#pdf("4-Clustering of module eigengene_MCI.pdf", height=6, width=9)

pdf(file.path(outputDir,paste0("Clustering of module eigengene (",current_area,"_",current_groupname,")_sample.pdf")), height=10, width=15)
plotDendroAndColors(geneTree, cbind( mergedColors),
                    c( "Module Color"),dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
dev.off()

```

### 4.3 New Merged Modules
```{r}
nGenes = length(gene_symbol)
# toppology overlap heatmap：
nSelect = 400 
# For reproducibility, we set the random seed 
set.seed(10); 
select = sample(nGenes, size = nSelect); 
selectTOM = dissTOM[select, select]; 
# There's no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster. 
selectTree = hclust(as.dist(selectTOM), method = clusterMethod)#average 
selectColors = mergedColors[select]; 
# Open a graphical window 
sizeGrWindow(9,9) 
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing 
# the color palette; setting the diagonal to NA also improves the clarity of the plot 
plotDiss = selectTOM^softPower; 
diag(plotDiss) = NA; 
{
    pdf(file.path(outputDir,paste0("Gene dendrogram and heatmap after merge (",current_area,"_",current_groupname,").pdf")), height=11, width=10)
    TOMplot(plotDiss, 
        selectTree, 
        selectColors, 
        main = "Network heatmap plot, selected genes") 
    dev.off()
}

```


### 4.4 Eigen values for merged modules
```{r}
MEList_merge = moduleEigengenes((as.matrix(data_WGCNA)), colors = mergedColors)
MEs = MEList_merge$eigengenes
# 计算根据模块特征向量基因计算模块相异度：
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average")
#METree = flashClust(as.dist(MEDiss), method = "average")
# Plot the result

plotEigengeneNetworks(MEs, 
                      "Eigengene adjacency heatmap", 
                      marHeatmap = c(3,4,2,2), 
                      plotDendrograms = FALSE, 
                      xLabelsAngle = 90) 
```

### 4.5 Module function annotations

```{r}
unique(mergedColors)


colors_mod <- unique(dynamicColors)
df_mod <- matrix(data=NA,ncol=length(colors_mod),nrow=max(table(dynamicColors)))
mf_unmerge_dir <- file.path(wdir,"moduleFunctions_unmerge")
mf_merge_dir <- file.path(wdir,"moduleFunctions_merge")
if (!dir.exists(mf_unmerge_dir)) {dir.create(mf_unmerge_dir)}#wdir
if (!dir.exists(mf_merge_dir)) {dir.create(mf_merge_dir)}


for(i in 1:length(colors_mod)){
  col_tmp <- colors_mod[i]
  genes_tmp <- colnames(data_WGCNA)[dynamicColors == col_tmp]
  
  gene_enrichment <- enricher(gene=genes_tmp, TERM2GENE = wp2gene_GO)
  #head(gene_enrichment)
  print(paste0("current color: ",col_tmp))
  print(head(gene_enrichment[,c("ID","p.adjust","Count","geneID")]))
  write.csv(x=gene_enrichment,file=paste0(mf_unmerge_dir,"/",col_tmp,"_signed_unmerged_functionAnnotation",current_area,"_",current_groupname,".csv"), row.names = FALSE)
  df_mod[(1:length(genes_tmp)),i] <- genes_tmp
}
colnames(df_mod) <- colors_mod
df_mod <- as.data.frame(df_mod)
write.csv(df_mod,file=paste0("ROSMAP_",current_area,"_",current_groupname,"_module_unmerged_signed.csv"))

colors_mod <- unique(mergedColors)
df_mod_merge <- matrix(data=NA,ncol=length(colors_mod),nrow=max(table(mergedColors)))
for(i in 1:length(colors_mod)){
  col_tmp <- colors_mod[i]
  genes_tmp <- colnames(data_WGCNA)[mergedColors == col_tmp]
  gene_enrichment <- enricher(gene=genes_tmp, TERM2GENE = wp2gene_GO)
  print(paste0("current color: ",col_tmp))
  print(head(gene_enrichment[,c("ID","p.adjust","Count","geneID")]))
  write.csv(x=gene_enrichment,file=paste0(mf_merge_dir,"/",col_tmp,"_signed_merged_functionAnnotation",current_area,"_",current_groupname,".csv"), row.names = FALSE)
  df_mod_merge[(1:length(genes_tmp)),i] <- genes_tmp
}
colnames(df_mod_merge) <- colors_mod
df_mod_merge <- as.data.frame(df_mod_merge)
write.csv(df_mod_merge,file=paste0("ROSMAP_",current_area,"_",current_groupname,"_module_merge.csv"), row.names = FALSE)
```

## Merged MEtree
```{r}

plot(METree, 
     main = "Clustering of module eigengenes",
     xlab = "", 
     sub = "")
# 在聚类图中画出剪切线
abline(h=MEDissThres, col = "red")
```

```{r}
table(mergedColors)
```

### 4.6 Output final module Summaries
```{r}
#rownames(MEs)<-paste0("ROSMAP_NCI",rownames(data_ROSMAPmt_NCI2))
write.table(MEs, file=paste0("ROSMAP_ME_output_",current_area,"_",current_groupname,".csv"),sep=",")
#write out final modules
write.table(table(mergedColors), file=paste0("ROSMAP_modules",current_area,"_",current_groupname,".csv"), sep=",")

```











